New modal operators for constructive mathematics

### 2023-09-11 slides-herrsching2023.tex
Beginning of Herrsching slides

{
\definecolor{mypurple}{RGB}{150,0,255}
\setbeamercolor{structure}{fg=mypurple}
\begin{frame}{}
  \begin{center}\includegraphics[width=0.2\textwidth]{eigenvector}\end{center}

@@

  \begin{columns}[c]
    \begin{column}{0.01\textwidth}
      \includegraphics[height=2.4em]{question-mark}
    \end{column}
    \begin{column}{0.9\textwidth}
      \mbox{Can locally the functions~$\lambda_i$ be chosen to be continuous?
      \only<2->{\hil{Yes.}}} \\
      How about the~$v_i$? \only<2->{\hil{No.}}
    \end{column}
  \end{columns}
\end{frame}
}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight]{night-sky}\end{minipage}}
\begin{frame}[c]
  \centering
  \color{white}

  \setbeamercolor{block body}{bg=black!100}
  \begin{minipage}{0.8\textwidth}
    \begin{block}{}
      \centering\normalsize\color{white}
      \hil{New modal operators for~constructive~mathematics}
    \end{block}
  \end{minipage}

  \emph{Type Theory, Constructive Mathematics and Geometric Logic} \\
  \bigskip

  CIRM \\
  May 2nd, 2023
  \bigskip

  Ingo Blechschmidt \\
  j.w.w. Alexander Oldenziel
  \par
\end{frame}}

\section{Motivation}

\begin{frame}{Questions}
  \begin{enumerate}
    \item \!Why has the inductive revolution been so powerful?
    \bigskip

    \item \!Why do proofs using Zorn's maximal ideals work so well in constructive algebra?
    \bigskip

    \item \!Why are elements of~$\bigcap_\ppp \ppp$ not necessarily nilpotent?
    \bigskip

    \item How can we extract computational content from classical proofs?
  \end{enumerate}
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight]{sea-of-clouds-2}\end{minipage}}
\begin{frame}{Infinite data}
  \vspace*{-1em}
  \[ \astikznodetransparentlycircled{xm}{7}\!,
    \quad \astikznodetransparentlycircled{x0}{4}\!,
    \quad \only<1-2>{\astikznodetransparentlycircled{t1}{3}}\only<3->{\astikznodecircled{t1}{mypurple}{3}}\!,
    \quad \only<1>{\ldots}\pause \astikznodetransparentlycircled{x1}{1}\!,
    \quad \only<2>{\ldots}\pause \astikznodecircled{t2}{mypurple}{8}\!,
    \quad \only<3>{\ldots} \visible<4->{\astikznodetransparentlycircled{x2}{2}\!,}
    \quad \only<4->{\ldots} \]
  {\centering\begin{tikzpicture}[remember picture,overlay]
    \node[draw=mypurple, circle, thick, inner sep=0.1em] (t3) {\scriptsize$\leq$};
    \path[draw=mypurple,thick]
      (t1)
      to [out=-90, in=180] (t3)
      to [out=0, in=-90] (t2);
  \end{tikzpicture}\par}
  \medskip
  \pause

  \textbf{Thm.} Every sequence~$\alpha : \NN \to \NN$ is \hil{good} in that
  there exist~$i < j$ with~$\alpha(i) \leq \alpha(j)$.
  \pause

  \emph{Proof.} \emph{(offensive?)} By~\badbox{\textsc{lem}}, there is a
  minimum~$\alpha(i)$.
  Set~$j \defeq i + 1$. \qed\par
  \pause
  \medskip

  \textbf{Def.} A preorder~$X$ is \hil{well\only<9->{$^\star$}} iff every sequence~$\NN \to X$ is good.

  \textbf{Examples.} $(\NN,{\leq}),\ \
  \color{white}\only<7->{\color{red!90}}\astikznode{onlyclass}{$\underbrace{\color{black}X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X)}_{\text{\visible<7->{\bad{only classically}}}}$}$.
  \pause
  \pause
  \medskip

  \begin{tikzpicture}[remember picture,overlay]
    \node[thick, fill=black, rectangle, inner sep=0.3em, right=2em of onlyclass] (moral) {
      \begin{minipage}{6cm}
        \begin{columns}
          \begin{column}{0.15\textwidth}
            \hspace*{1.0em}\color{white}\dbend
          \end{column}
          \begin{column}{0.95\textwidth}
            \color{white}\footnotesize
            \it Don't quantify over points of spaces which might not have enough.
          \end{column}
        \end{columns}
      \end{minipage}
    };
    \path[draw=red!90,thick,-stealth]
      (moral) to
      [out=180, in=-00] (onlyclass.east);
  \end{tikzpicture}
  \pause

  \textbf{Def.} For a predicate~$P$ on finite lists over a set~$X$, inductively
  define:
  \[
    \infer{P \,|\, \sigma}{P\sigma}
    \qquad
    \infer{P \,|\, \sigma}{\forall(x \in X)\_\ P \,|\, \sigma x}
  \]

  \textbf{Def.} A preorder is \hil{well} iff~$\mathsf{Good} \,|\, [\,]$, where
  $\mathsf{Good}\,\sigma \defeqv (\exists(i < j)\_ \sigma[i] \leq \sigma[j])$.
\end{frame}}

\begin{frame}{Computational content from classical proofs}
  \textbf{Def.} A transitive relation~$({<})$ on a set~$X$ is \ldots
  \begin{enumerate}
    \item \hil{well-founded$^\star$} iff
    there is no \bad{infinite chain} $x_0 > x_1 > \cdots$,
    \item \hil{well-founded} iff for every~$x \in X$,
    $\mathsf{Acc}(x)$,
  \end{enumerate}
  where~$\mathsf{Acc}$ is inductively defined by:
  \[
    \infer{\mathsf{Acc}(x)}{\forall(y < x)\_\ \mathsf{Acc}(y)}
  \]
  \vspace*{-1em}
  \pause

  \textbf{Prop.} Let~$(X,{\leq})$ be preorder. Let~``$x < y$'' mean~$x \leq y \wedge \neg(y \leq x)$. \\
  Then: If~$X$ is well$^\star$, then~$({<})$ is well-founded$^\star$.

  \emph{Proof.} An infinite strictly descending chain would also be good. \qed
  \pause

  Can we extract a constructive proof that well preorders are well-founded?
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
\begin{frame}{Maximal ideals}
  \only<1-7>{\textbf{Thm.}
  Let~$M$ be a surjective matrix with more rows than columns over a
  ring~$A$. Then~$1 = 0$ in~$A$.

  \visible<2->{\emph{Proof.} \bad{Assume not.}}
  \visible<3->{Then there is~a \bad{maximal ideal} $\mmm$.}
  \visible<5->{The matrix is surjective over~$A/\mmm$.}
  \visible<6->{Since~$A/\mmm$ is a field, this is a contradiction to basic linear algebra.\qed}}

  \only<4-7>{\bigskip\par\centering\scalebox{0.9}{\centering\begin{tikzpicture}
    \node (0) at (0,1) {$(0) = \{0\}$};
    \node (1) at (0,5) {$(1) = \ZZ$};
    \node (2) at (-2,4) {$(2)$};
    \node [right of=2] (3) {$(3)$};
    \node [below of=2] (4) {$(4)$};
    \node [below of=2, xshift=0.7cm] (6) {$(6)$};
    \node [right of=3] (5) {$(5)$};
    \node [right of=5] (7) {$(7)$};
    \node [right of=7] (7d) {$\ldots$\phantom{(}};
    \node [right of=7d, xshift=3cm, yshift=-2cm] (max)
    {\vbox{\small\it maximal among the proper ideals \\ \medskip $\neg(1 \in
    \mmm)$ \\ \medskip $x \in \mmm \quad\vee\quad 1 \in \mmm + (x)$}};
    \node [below of=4] (8) {$(8)$};
    \node [right of=8, xshift=3cm] (8d) {$\ldots$};
    \draw (0) -- (8);
    \draw (0) -- (8d);
    \draw (0) -- (6);
    \draw (2) -- (1);
    \draw (3) -- (1);
    \draw (5) -- (1);
    \draw (7) -- (1);
    \draw (7d) -- (1);
    \draw (4) -- (2);
    \draw (8) -- (4);
    \draw (6) -- (2);
    \draw (6) -- (3);
    \draw [mypurple!30, thick, shorten <=-2pt, shorten >=-2pt, ->] (max) to [out=120, in=-30] (7d);
    \begin{pgfonlayer}{background}
      \draw[decorate, very thick, draw=mypurple!30]
        ($(2.south west) + (8pt, 0)$) arc(270:180:8pt) --
        ($(2.north west) + (0, -8pt)$) arc(180:90:8pt) --
        ($(7d.north east) + (-8pt, 0)$) arc(90:0:8pt) --
        ($(7d.south east) + (0, 8pt)$) arc(0:-90:8pt) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}\par}\par}

  \pause
  \pause
  \pause
  \pause
  \pause
  \pause
  \medskip
  \raggedright

  Let~$A$ be a ring. \emph{Does there exist a maximal ideal~$\mmm \subseteq A$?}
  \pause
  \begin{enumerate}
    \item \good{Yes}, if \bad{Zorn's lemma} is available.
    \bigskip
    \pause

    \item \good{Yes}, if~$A$ is countable and membership of finitely generated ideals is decidable:
    {\footnotesize
    \begin{align*}
      \mmm_0 &\defeq \{ 0 \}, &
      \mmm_{n+1} &\defeq \begin{cases}
        \mmm_n + (x_n), & \text{if $1 \not\in \mmm_n + (x_n)$}, {\qquad\quad\quad\!\!}\\
        \mmm_n, & \text{else.}
      \end{cases}
    \end{align*}}
    \pause

    \item \good{Yes}, if~$A$ is countable (irrespective of membership decidability):
    {\footnotesize\begin{align*}
      \mmm_0 &\defeq \{ 0 \}, &
      \mmm_{n+1} &\defeq \mmm_n + (\underbrace{\{ x \in A \,|\, x = x_n \wedge
        1 \not\in \mmm_n + (x_n) \}}_{\text{a certain subsingleton set}})
    \end{align*}}
    \vspace*{3.5em}
    \visible<11>{{
      \centering
      \raisebox{0pt}[0pt][0pt]{
        \hspace*{1.5em}
        \begin{tikzpicture}
          \node (inner) at (17.3mm,-20mm) {\textit{``a bad joke''}};
          \path (0,0) pic{laurel-wreath};
        \end{tikzpicture}
        \hspace*{-3em}
        \begin{tikzpicture}
          \node (inner) at (17.5mm,-18mm) {\vbox{\small\centering\textit{``non- \\informative''}}};
          \path (0,0) pic{laurel-wreath};
        \end{tikzpicture}
      }
      \par
    }}
    \pause
    \pause
    \vspace*{-3.5em}

    \item In the general case: \bad{No}\pause, \\ but \emph{first-order
    consequences} of the existence of a maximal ideal \good{do hold}.
  \end{enumerate}
\end{frame}}

\begin{frame}{Questions}
  \begin{enumerate}
    \item \!Why has the inductive revolution been so powerful?
    \bigskip

    \item \!Why do proofs using Zorn's maximal ideals work so well in constructive algebra?
    \bigskip

    \item \!Why are elements of~$\bigcap_\ppp \ppp$ not necessarily nilpotent?
    \bigskip

    \item How can we extract computational content from classical proofs?
  \end{enumerate}
\end{frame}

\renewcommand{\insertframeextra}{a}
{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.23cm}\includegraphics[width=\paperwidth]{topos-horses-lighter-lighter}\end{minipage}}
\begin{frame}{Constructive forcing (= Grothendieck toposes)}
  \justifying
  Let~$L$ be a \hil{forcing notion}, a preorder equipped
  with a \hil{covering system}.\only<1-2>{\footnote{\scriptsize A covering system consists of a
  set~$\Cov(\sigma) \subseteq P({\downarrow}\sigma)$ of \emph{coverings} for
  each element~$\sigma \in L$ subject only to the following simulation
  condition: If~$\tau \preccurlyeq \sigma$ and~$R \in \Cov(\sigma)$, there
  should be a covering~$S \in \Cov(\tau)$ such that~$S \subseteq
  {\downarrow}R$.}}\only<3->{\textsuperscript{\phantom{1}}}
  \hil{Filters}~$F \subseteq L$ are subsets which are upward-closed,
  downward-directed and split the covering system.\only<1-2>{\footnote{\scriptsize
  A subset~$F$ splits the covering system iff for every~$\sigma \in L$ and~$R \in \Cov(\sigma)$,
  if~$\sigma \in F$, then~$\tau \in F$ for some~$\tau \in R$.}}\only<3>{\textsuperscript{\phantom{2}}}
  \pause

  {\centering\footnotesize\begin{tabular}{llll}
    \toprule
    & preorder~$L$ & coverings of an element~$\sigma \in L$ & filters of~$L$ \\
    \midrule
    \normalnumber{1} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$ & maps~$\NN \to X$ \\
    \normalnumber{2} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$,\ \ $\{ \sigma\tau \,|\, \tau \in X^*, a \in \sigma\tau \}$ for each~$a \in X$ & surjections~$\NN \twoheadrightarrow X$ \\
    \normalnumber{3} & f.g. ideals & --- & ideals \\
    \normalnumber{4} & f.g. ideals & $\{ \sigma+(a), \sigma+(b) \}$ for each~$ab \in \sigma$,\ \ $\{\}$ if~$1 \in \sigma$ & prime ideals \\
    \normalnumber{5} & opens & $\mathcal{U}$ such that~$\sigma = \bigcup \mathcal{U}$ & points \\
    \normalnumber{6} & $\{\star\}$ & $\{ \star \,|\, \varphi \} \cup \{ \star \,|\, \neg\varphi \}$ &
    witnesses of~\textsc{lem}
    \\
    \bottomrule
  \end{tabular}\par}
  \bigskip

  \only<3>{\textbf{Def.} Given a monotone predicate~$P$
  on~$L$, inductively define:
  \[
    \infer{P \,|\, \sigma}{P\sigma}
    \qquad
    \infer[{\small(R \in \Cov(\sigma))}]{P \,|\, \sigma}{\forall(\tau \in R)\_\ P \,|\, \tau}
  \]
  We use quantifier-like notation: ``$\nabla\sigma\_ P\sigma$'' means~$P \,|\, \sigma$.}
\end{frame}}

\addtocounter{framenumber}{-1}
\renewcommand{\insertframeextra}{b}
{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.23cm}\includegraphics[width=\paperwidth]{topos-horses-lighter-lighter}\end{minipage}}
\begin{frame}{Constructive forcing (= Grothendieck toposes)}
  A forcing notion is a template for a \hil{forcing extension}~$V^\nabla$
  of the base universe~$V$:
  \medskip

  \begin{center}\begin{tikzpicture}[remember picture,overlay]
    \node[thick, draw=mypurpledark, rectangle, inner sep=0.3em] (kj) {
      \begin{minipage}{0.6\textwidth}
        \centering
        \emph{When we say that a statement \hil{holds in~$V^\nabla$},} \\
        we mean that its~$\nabla$-translation is true.\par
        \vspace*{0.3em}
      \end{minipage}
    };
  \end{tikzpicture}\end{center}
  \bigskip
  \pause

  \textbf{Examples.}
  \begin{enumerate}
  \justifying
  \item For~$L = \{\star\}$, ``every number is zero or a successor'' holds
  in~$V^\nabla$ because

  \vspace*{-1em}
  \[ \forall(n \in \NN)\_ \neg\neg\bigl(\neg\neg(n = 0) \vee
    \neg\neg(\exists(m \in \NN)\_ \neg\neg(n = S(m)))\bigr). \]

  \pause
  \item For~$L$ associated to a space, ``every symmetric matrix has an
  eigenvector'' holds in~$V^\nabla$ iff for every continuous family of
  symmetric matrices, locally there is a continuous eigenvector-picking
  function.
  \pause
  \medskip

  \item For any~$L$, ``there is a filter of~$L$'' holds in~$V^\nabla$,
  witnessed by the \hil{generic filter} of~$L$.
  \pause
  \medskip

  \item Let~$X$ be a preorder. Then:

  \hil{$\blacktriangleright$} $X$ is well iff the generic sequence~$\NN \to X$
  is good.

  \hil{$\blacktriangleright$} $({<})$ is well-founded iff the generic strictly
  descending chain validates~$\bot$.
  \end{enumerate}
\end{frame}}

\renewcommand{\insertframeextra}{}
\begin{frame}{The modal multiverse}
  \only<1-2>{In general, ``$\varphi$ holds in~$V^\nabla$'' and~``$\varphi$ holds in~$V$''
  are \emph{not} equivalent.

  \hil{$\blacktriangleright$}
  For \hil{positive} extensions, they are equivalent for coherent implications.\\
  %\footnote{bounded first-order formulas of the
  %form~$\forall\ldots\forall\_(\ldots\Rightarrow\ldots)$,
  %where antecedent and succedent do not contain~``$\forall$''
  %and~``${\Rightarrow}$''}.
  \phantom{\hil{$\blacktriangleright$}} {\footnotesize -- e.g. the ``Barr cover''.\par}

  \hil{$\blacktriangleright$}
  For \hil{positive overt} extensions, they are equivalent for bounded first-order formulas. \\
  \phantom{\hil{$\blacktriangleright$}} {\footnotesize -- e.g. $V^\nabla$ containing
  the generic surjection~$\NN \twoheadrightarrow X$, if~$X$ is inhabited.\par}
  \bigskip}
  \pause

  \textbf{Def.} A statement~$\varphi$ holds \ldots
  \begin{enumerate}
    \small
    \item[$\blacktriangleright$] \hil{everywhere} ($\necessary\varphi$) iff
    it holds in every extension.
    \item[$\blacktriangleright$] \hil{somewhere} ($\possible\varphi$) iff
    it holds in some positive extension.
    \item[$\blacktriangleright$] \hil{proximally} ($\xpossible\varphi$) iff
    it holds in some positive overt extension.
  \end{enumerate}
  \bigskip

  \visible<2>{\begin{tikzpicture}[overlay]
    \node[anchor=south east,inner sep=0] (image) at (14.8,-0.8) {
      \includegraphics[width=0.2\textwidth]{branching}
    };
  \end{tikzpicture}\vspace*{-1.6em}}
  \visible<3>{\begin{tikzpicture}[overlay]
    \node[anchor=south east,inner sep=0] (image) at (14.8,0.9) {
      \includegraphics[width=0.2\textwidth]{multiverse}
    };
  \end{tikzpicture}\vspace*{-0.3em}}

  \only<2>{\scriptsize Foreshadowed by:
  \begin{itemize}
    \scriptsize
    \item[1984\phantom{s}] André Joyal, Miles Tierney. \emph{An extension of the Galois theory of
    Grothendieck.}
    \item[1987\phantom{s}] Andreas Blass. \emph{Well-ordering and induction in intuitionistic
    logic and topoi.}
    \item[2010s] Milly Maietti, Steve Vickers. Ongoing work on arithmetic
    universes.
    \item[2011\phantom{s}] Joel David Hamkins. \emph{The set-theoretic multiverse.}
    \item[2013\phantom{s}] Shawn Henry. \emph{Classifying topoi and preservation of higher
    order logic by geometric morphisms.}
  \end{itemize}}
  \pause

  \small
  \begin{multicols}{2}
    For every inhabited set~$X$, \emph{proximally} \\
    there is an enumeration~$\NN \twoheadrightarrow X$.
    \bigskip

    A preorder is well iff \emph{everywhere}, \\
    every sequence is good.
    \bigskip

    A ring element is nilpotent iff \\
    all prime ideals \emph{everywhere} contain it.
    \bigskip

    For every ring, \emph{proximally} \\ there is a maximal ideal.
    \bigskip

    A relation is well-founded iff \emph{everywhere}, there is no descending chain.
    \bigskip

    \emph{Somewhere}, \\ the law of excluded middle holds.
  \end{multicols}
\end{frame}

\begin{frame}{The multiverse perspective}
  \begin{enumerate}
    \item \!\emph{Why has the inductive revolution been so powerful?}

    Because the inductive conditions
    are equivalent to truth in \emph{all} forcing extensions.
    \bigskip

    \item \!\emph{Why do proofs using Zorn's maximal ideals work so well in constructive algebra?}

    Because every ring proximally has a maximal ideal.
    \bigskip

    \item \!\emph{Why are elements of~$\bigcap_\ppp \ppp$ not necessarily nilpotent?}

    Because we forgot the prime ideals in forcing extensions.
    \bigskip

    \item \emph{How can we extract computational content from classical proofs?}

    By traveling the multiverse (upwards, keeping ties to the base), exploiting
    that \\
    -- \textsc{lem} holds \emph{somewhere} and \\
    -- \textsc{dc} holds \emph{proximally}.
  \end{enumerate}
\end{frame}

\backupstart

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{eventually-monad-cropped}\end{minipage}}
\begin{frame}
  \vspace*{\paperheight}\vspace*{-2.05em}\centering
  \emph{Partial Agda formalization available.}
\end{frame}}

\backupend

@@

>>>>

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[height=\paperheight]{multiverse-faded}\end{minipage}}
\begin{frame}[c]
  \centering
  \color{white}

  \setbeamercolor{block body}{bg=black!100}
  \begin{minipage}{0.35\textwidth}
    \begin{block}{}
      \centering\normalsize\color{white}
      \hil{Constructive forcing}
    \end{block}
  \end{minipage}

%  \colorbox{black!50}{Autumn school on} \\
%  \colorbox{black!50}{\emph{Proof and Computation}} \\
%  \colorbox{black!50}{in Herrsching} \\
%  \ \\
%  \colorbox{black!50}{September 20th to September 16th, 2023}
%
%  \colorbox{black!50}{Ingo Blechschmidt} \\
%  \par
\end{frame}}

\begin{frame}{}
  \begin{center}\includegraphics[width=0.2\textwidth]{eigenvector}\end{center}

@@

  \begin{columns}[c]
    \begin{column}{0.01\textwidth}
      \includegraphics[height=2.4em]{question-mark}
    \end{column}
    \begin{column}{0.9\textwidth}
      \mbox{Can locally the functions~$\lambda_i$ be chosen to be continuous?
      \only<2->{\hil{Yes.}}} \\
      How about the~$v_i$? \only<2->{\hil{No.}}
    \end{column}
  \end{columns}
\end{frame}

% \begin{document}

\newcommand\ytl[2]{%
  \parbox[b]{4.5em}{\hfill{#1}~$\cdots\cdots$~}%
  \makebox[0pt][c]{\color{mypurple}$\bullet$}{\color{mypurple}\vrule}\quad%
  \parbox[c]{9.3cm}{\vspace{7pt}\raggedright#2\\[7pt]}%
  \\[-2.0pt]}
\begin{frame}{A brief timeline}
  \begin{columns}[t]
    \begin{column}{0.86\textwidth}
      \vspace*{-1.0em}
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-3.5pt]
      \ytl{1878}{Cantor advances the \hil{continuum hypothesis}, the claim
      that~$2^{\aleph_0} = \aleph_1$.}
      \pause
      \ytl{1910s}{Zermelo--Fraenkel set theory emerges.}
      \pause
      \ytl{1920s}{Set theorists pursue additional axioms to
      settle~\textsc{ch} \\ (one way or another).}
      \pause
      \ytl{1938}{Gödel proves: If~\textsc{zfc} is consistent, so
      is~\textsc{zfc}+\textsc{ch}.}
      \pause
      \ytl{1963}{Cohen proves: If~\textsc{zfc} is consistent, so
      is~\textsc{zfc}+$\neg$\textsc{ch}.}
      \pause
      \ytl{2011}{Hamkins offers his paper on the \hil{multiverse position} in
      the philosophy of set theory.}
      \pause
      %arguing that the program of pursuing
      %additional axioms (while successful in many ways) is doomed to fail
      %to settle~\textsc{ch}.}
      \ytl{2016}{Oldenziel proposes to study the modal multiverse of parametrized
      mathematics.}
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
    \end{column}

    \begin{column}{0.20\textwidth}
      \pause
      \centering\includegraphics[width=\textwidth,valign=t]{roitman}

      \scriptsize Judith Roitman %(* 1945)
      \medskip

      \emph{Mainstream mathematics is beginning to see results using
      modern set theoretic techniques.}
    \end{column}
  \end{columns}
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
\begin{frame}{Constructive forcing}
  \medskip
  \begin{columns}[t]
    \begin{column}{0.40\textwidth}
      \quad\includegraphics[width=0.8\textwidth,valign=t]{branching}
    \end{column}

    \begin{column}{0.75\textwidth}
      \hil{Extending the universe} in various ways, \\[0.7em]
      \pause
      \quad similarly how we can extend groups or rings, \\[0.7em]
      \pause
      \quad\quad in and for \hil{constructive mathematics} \\[0.7em]
      \pause
      \quad\quad\quad \hil{without} presupposing familiarity with \\[0.7em]
      \pause
      \quad\quad\quad\quad set theory, topos theory, or sheaves.
      \pause
    \end{column}
  \end{columns}

  \bigskip
  \bigskip

  \emph{Outline:}
  \vspace*{-0.5em}
  \begin{enumerate}
    \item What can forcing do for you?
    \item Forcing notions and Kripke--Joyal semantics
    \item Case studies in constructive algebra and combinatorics
  \end{enumerate}
\end{frame}}

@@

- Judith Roitman
- Proof and Computation

- What can forcing do for you?
  - Exploring the range of foundational possibility: CH, ¬CH, FTA, ...
  - Eigenvalues
  - Generic sequence
  - Generic surjection
  - Matthias
  - "Harness the flexibility of parametric mathematics in an accessible way."

Overloading

1. Eigenvalues
2. Generic surjection
3. Generic sequence
4. FTA

Extending the universe in various ways,

similarly how we can extend groups or rings,

in the context of and for the purposes of constructive mathematics

without requiring knowledge of

set theory, topos theory, category theory, sheaves.
